Better fusion for filters

### 2014-04-12 2014betterfusionforfilters/Main.tex
Wibble
explicitly state similarities and differences to existing ILP formulations
Add diagrams for normalize2 example
size inference

% -----------------------------------------------------------------------------
\begin{document}
\title{Better fusion for filters}

\authorinfo{ 
  Amos Robinson$^\dagger$ 
  \and Ben Lippmeier$^\dagger$
  \and Manuel M. T. Chakravarty$^\dagger$
  \and Gabriele Keller$^\dagger$ 
}{
  \vspace{5pt}
  \shortstack{
    $^\dagger$Computer Science and Engineering \\
    University of New South Wales, Australia \\[2pt]
    \textsf{\{amosr,benl,chak,keller\}@cse.unsw.edu.au}
  }
}

@@

\begin{abstract}
Recent work on array fusion demonstrates how to compile a specific class of data flow programs into single imperative loops. However, until now there has been no satisfactory way to take the operator graph expressing a complete program and split it into fusible subgraphs. Finding good clusterings of operators is non-trivial, as we must achieve a balance between minimizing memory traffic and loop overhead, while maximizing cache locality. We show how to formulate operator clustering as an \emph{integer linear programming} problem, and use an external library to find an good solution. 
\end{abstract}

% BL: I've removed uses of the word "optimal" here because we don't have a proper
% language cost model, and our solution is ultimately based on heuristics. ILP can find 
% an "optimal" solution for a heuristic cost metric, but this isn't the thing we're
% really trying to optimize.

@@

>>>>

% -----------------------------------------------------------------------------
\begin{document}
\title{Better fusion for filters}
\authorinfo{ 
  Amos Robinson$^\dagger$ 
  \and Ben Lippmeier$^\dagger$
  \and Manuel M. T. Chakravarty$^\dagger$
  \and Gabriele Keller$^\dagger$ 
}{
  \vspace{5pt}
  \shortstack{
    $^\dagger$Computer Science and Engineering \\
    University of New South Wales, Australia \\[2pt]
    \textsf{\{amosr,benl,chak,keller\}@cse.unsw.edu.au}
  }
}

@@

\begin{abstract}
Recent work on array fusion demonstrates how to compile a specific class of data flow programs into single imperative loops.
However, until now there has been no satisfactory way to take the operator graph expressing a complete program and split it into fusible subgraphs.
Finding good clusterings of operators is non-trivial, as we must achieve a balance between minimizing memory traffic and loop overhead, while maximizing cache locality.
We show how to formulate operator clustering as an \emph{integer linear programming} problem, and use an external library to find a good solution. 
\end{abstract}

% BL: I've removed uses of the word "optimal" here because we don't have a proper language cost model, and our solution is ultimately based on heuristics. ILP can find  an "optimal" solution for a heuristic cost metric, but this isn't the thing we're really trying to optimize.

@@

\section*{Acknowledgements}
Many thanks are due to
Robert Clifton-Everest,
Kai Engelhardt,
Bill Kroon,
Frederik Madsen,
Abdallah Saffidine,
and Jingling Xue
for enlightening discussions relating to this work.

### 2014-04-12 2014betterfusionforfilters/section/03-SizeInference.tex
size inference

After size inference, the goal is to have a function that only requires runtime size checks at the start of the function. This outlaws certain filter constraints.

@@

Talk about the relation to Chatterjee\cite{chatterjee1991size}.

@@

\begin{tabbing}
MMMMM       \= MM \= MMMM \= MM \= \kill
$constrs$  \> @::@  \> $bind$  \> $\to$ \> $\{var = rate\}$ \\
MMMMMMMMMMMMMM               \= MM  \= \kill
$constrs(o = @fold@~f~n)  $ \> $=$ \> $\{\}$     \\
$constrs(o = @map@_n~f~ns)$ \> $=$ \> $\{o = n~|~n \in ns\}$     \\
$constrs(o = @filter@~f~b)$ \> $=$ \> $\{o = Filtered_b\}$              \\
$constrs(o = @gather@~i~d)$ \> $=$ \> $\{o = i\}$ \\
$constrs(o = @cross@~a~b) $ \> $=$ \> $\{o = a \times b\}$ \\
$constrs(outs = @external@~ins) $ \> $=$ \> $\{o = External_o~|~o \in outs\}$ \\
\end{tabbing}

Then, for some program $\lambda\ldots~@let@~binds~@in@~\ldots$, find the union of all constraints:
$\bigcup_{b \in binds} constrs(b)$.
The generated equalities are grouped into equivalence classes.

All combinators whose constraints are in the same equivalence class have the same iteration size, and may be fused together.
These constraints are also used to rule out any programs that require runtime checks on array lengths during the program.
In other words, only programs that can be statically determined to require only array length checks at the start of execution.

\TODO{Well\ldots} note that $Filtered_n$, $External_n$ are \emph{existential} constructors, bound at $n$.
This basically translates into them not being able to be referred to \emph{before} their bindings.
Also note that there can be at most one distinct constructor per equivalence class.
Go through the examples below.

\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 us                   {us   = k_us}
  = let sum1 = fold   (+) 0 us   
        gts  = filter (>0)  us   {gts  = Filtered_us}
        sum2 = fold   (+) 0 gts   
        nor1 = map  (/sum1) us   {nor1 = k_us}
        nor2 = map  (/sum2) us   {nor2 = k_us}
    in (nor1, nor2)
\end{code}

\begin{code}
 { us   : us = nor1 = nor2
 , gts  : Filtered_us }
\end{code}

\subsection{Disallowed filters}

Filtered constraints of the form $a \le_c b$ are special.
The actual size of the filtered output is unknown, and so cannot be constrained.
For example, consider the constraints generated by this program:
\begin{code}
test vs    =
 let flt   = filter p vs
     wrong = map2   f flt vs
 in  wrong
\end{code}
With the resulting constraints being $\{flt \le_c vs, wrong =_c flt, wrong =_c vs\}$.
By transitivity, we find that $flt \le_c flt$ - the size of the filtered output is upper-bounded by the size of the filtered output! 
Why is this bad?
Actually implementing this would require to buffer the entire contents of @flt@, so for simplicity we disallow it. Talk more about this.

Similarly, this example:
\begin{code}
test vs    =
 let flt   = filter p  vs
     flt2  = filter p' vs
     wrong = map2   f  flt flt2
 in  wrong
\end{code}
Generates the constraints $\{flt \le_c vs, flt2 \le_c vs, flt =_c flt2\}$.
Here we are constraining two different filtered outputs to be the same size.
We disallow this.

Not all constraints on filtered data are disallowed, of course;
\begin{code}
test vs    =
 let flt   = filter p  vs
     xs    = map    f  flt
     ys    = map    g  flt
     sm    = map2   f  xs ys
 in  sm
\end{code}
This generates the constraints $\{flt \le_c vs, xs =_c flt, ys =_c flt, sm =_c xs, sm =_c ys\}$,
but this is perfectly valid, since the constraints are all actually based on @flt@.

>>>>

@@

Size inference has been explored before in the context of fusion by Chatterjee~\cite{chatterjee1991size}, but the only constraints they support are equality.
Their formulation has no filtering operations, and all array operations in their case generate the same size as at least one of the operation's inputs.
A more in-depth analysis has also been explored by Jay~\cite{jay1996shape} in the form of shape inference, which is built from primitives such as @cons@ and @nil@.
Since we have fewer and simpler primitives, our size inference does not need to be as complicated as Jay's shape inference.

@@

\begin{tabbing}
MMMMM       \= MM \= MMMM \= MM \= \kill
$constrs$  \> @::@  \> $bind$  \> $\to$ \> $\{var = rate\}$ \\
MMMMMMMMMMMMMM               \= MM  \= \kill
$constrs(o = @fold@~f~n)  $ \> $=$ \> $\{\}$     \\
$constrs(o = @map@_n~f~ns)$ \> $=$ \> $\{o = n~|~n \in ns\}$     \\
$constrs(o = @filter@~f~b)$ \> $=$ \> $\{o = Filtered_o\}$              \\
$constrs(o = @gather@~i~d)$ \> $=$ \> $\{o = i\}$ \\
$constrs(o = @cross@~a~b) $ \> $=$ \> $\{o = a \times b\}$ \\
$constrs(outs = @external@~ins) $ \> $=$ \> $\{o = External_o~|~o \in outs\}$ \\
\end{tabbing}

Then, for some program $\lambda\ldots~@let@~binds~@in@~\ldots$, find the union of all constraints:
$\bigcup_{b \in binds} constrs(b)$.
The constraints are all equalities, so the generated equalities are easily grouped into equivalence classes.

All combinators whose constraints are in the same equivalence class have the same iteration size, and may be fused together.
These constraints are also used to rule out any programs that require runtime checks on array lengths during the program.

\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 us                   {us   = k_us}
  = let sum1 = fold   (+) 0 us   
        gts  = filter (>0)  us   {gts  = Filtered_gts}
        sum2 = fold   (+) 0 gts   
        nor1 = map  (/sum1) us   {nor1 = k_us}
        nor2 = map  (/sum2) us   {nor2 = k_us}
    in (nor1, nor2)
\end{code}

\subsection{Runtime checks on array lengths}
The $@map@_n$ combinator requires all its input arrays to be the same size.
We do not, however, want to perform runtime checks on the size of arrays before every combinator.
We have decided to only fuse programs that can be statically determined to require only array length checks at the start of execution.
Any other programs are simply left alone by our fusion system, potentially falling back to stream fusion.

We use the constraints to rule out any programs that require runtime checks on array lengths during the program.
Firstly, filter and external constraints are only allowed to be mentioned after their binding site, as it is only at that stage that the lengths are actually known.
Secondly, it is not possible to constrain one constructor to equal another: $Filtered_n = External_n$ is invalid.
Finally, any initial equality constraints, such as two input arrays being the same length, or one input array being the length of the cross product of two others, are dealt with by inserting runtime checks only at the start of the function.

The constraints $Filtered_n$, $External_n$ are treated as \emph{existential} constructors, bound at $n$.
This means that such constraints are not able to be referred to before their bindings.
Intuitively, we know that the output of a @filter@ or an @external@ has \emph{some} length, but we don't know what its value is until after they have executed.
Thus, constraining the input of something to be some exact length, before we know what the length is, is impossible.

Similarly, these constructors cannot be constrained to equal any other constructor. 
For example, $Filtered_n = External_n$ is invalid, as is $a \times b = Filtered_n$.
This corresponds to each equivalence class having \emph{at most} one distinct constructor associated with it.
Any other elements of the equivalence class must be either the same constructor, or simply rate variables.

\subsubsection{Examples}
Following are a few examples of disallowed programs.
Consider the constraints generated by this program:
\begin{code}
test vs    =
 let flt   = filter p vs        { flt   = Filtered_flt }
     wrong = map2   f flt vs    { wrong = flt = vs     }
 in  wrong
\end{code}
By transitivity, we find that $vs = Filtered_{flt}$, but $vs$ is bound before @flt@.
Operationally, actually implementing this would require to buffer the entire contents of @flt@, then check the length of @flt@ against the length of @vs@.

Similarly, this example:
\begin{code}
test vs    =
 let flt   = filter p  vs       { flt   = Filtered_flt  }
     flt2  = filter p' vs       { flt2  = Filtered_flt2 }
     wrong = map2   f  flt flt2 { wrong = flt = flt2    }
 in  wrong
\end{code}
Again, transitivity here requires that $Filtered_{flt} = Filtered_{flt2}$, but since these are both concrete constructors, we cannot have such a constraint.
Basically, we are constraining two different filtered outputs to be the same size.
Operationally, implementing this would require to buffer both @flt@ and @flt2@, before checking the two lengths.
